Recursive partitioning of linear growth curve models

### 2022-12-02 pkg/glmertree/paper/lgcm-tree/_Partitioning_GCMs_with_GLMM_trees.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/partykit@3240 255f4249-2263-4417-b1ea-8498779e27a7
Former-commit-id: 6fd67765c7e27182cd2bbfc253c1c292279af2b0

\abstract{Growth curve models are popular tools for studying {\color{red!80} the development of a response variable within the same observational units} over time. Often, researchers have a specific interest in uncovering subgroups that show different patterns of growth over time. Recursive partitioning methods (RPMs) allow for detecting such subgroups. In this paper, we focus on the use of generalized linear mixed effects model trees (GLMM trees) for partitioning linear growth curve models (LGCMs). GLMM trees were originally proposed for subgroup detection in clustered cross-sectional data. {\color{red!80} Here, we introduce an extension of the algorithm for longitudinal (as opposed to clustered) data that can detect subgroups defined by subject-specific covariates on/or time-varying covariates. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case.} Using simulated and empirical data, we compare the performance of GLMM trees and other methods for partitioning LGCMs: SEM trees, LongCART and longRPart. We find that GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for the analysis of trajectories with both discrete and continuous time, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

In longitudinal studies, (generalized) linear growth curve modeling is often used to model change over time, and inter-individual differences in change. For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail. Inter-individual heterogeneity is likely in such studies, where (groups of) participants show higher or lower rates of growth over time. Through the use of mixed-effects or latent-variable models, such heterogeneity can be captured, and explained by covariates of a-priori known relevance \citep{NeisyMatt18}.

Often, the covariates that predict or affect the trajectories may not be known a-priori, or researchers may have a specific interest in finding subgroups of participants with similar trajectories. Recursive partitioning methods (RPMs), also known as decision-tree methods, are particularly suited in such cases. Several existing RPMs allow for partitioning growth trajectories: For example, GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Partitioning based on non-linear growth curve models can be performed with the longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods\footnote{Both IT methods specifically focus at detecting so-called \textit{treatment-subgroup interactions}: subgroups that respond more favorably to one treatment than another.}.

In this paper, we focus on generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18}, which allow for subgroup detection in a wide range of mixed-effects models.  {\color{red!80} FIXME: The rest of this paragraph needs to be a bit more forceful!} Although originally not specifically developed for partitioning LGCMs, application of GLMM trees for partitioning LGCMs is straightforward. We propose two adaptations to the algorithm, which we expect to be of particular relevance for partitioning LGCMs.

The main difference between GLMM trees and the aforementioned RPMs, is that the latter all fit the full parametric model in each of the nodes. In contrast, GLMM trees uses the observations in each node to estimate the fixed-effects parameters, while using all observations to estimate the random-effects parameters. This may not only reduce computation time; it also reduces the total number of parameters to be estimated, which may improve stability and reduce overfitting. This local/global estimation approach was originally proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant means in the terminal nodes; \cite{FokkySmit18} generalized it to trees containing generalized linear models in the nodes, i.e., potentially including regressors and/or using a non-Gaussian family.

The longRPart, longRPart2, IT-LT and original SEM tree algorithms employ an exhaustive split detection procedure: For every possible cut point in the current node, the full parametric model is re-estimated in the two resulting nodes. To choose the optimal cutpoint, a likelihood ratio test is computed for every possible cut point. Drawbacks of this exhaustive search are a heavy computational burden, as well as a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}.

LongCART, MELT, GEE-based decision trees and score-based SEM trees fit full parametric models in each of the nodes as well, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces the computational burden. Furthermore, these approaches allow for employing a two-step approach for split detection, which separates variable and cutpoint selection, thus preventing variable selection bias. The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.

The remainder of this paper is structured as follows: In the next section (Estimation) we explain how GLMM trees are estimated, and how estimation can be adjusted to (better) account for dependence structures encountered in LGCMs. Next, in Study I, we perform a simulation study to assess performance of GLMM trees and the proposed adaptations in partitioning LGCMs. In Study II, we use the same simulation design to compare the performance of GLMM trees with that of two other mixed-effects RPMs: SEM trees and LongCART.  In Study III, we compare the performance of GLMM trees and longRPart in partitioning children's trajectories of reading, math and science abilities. In the Discussion, we summarize our findings and their implications.

@@

\section{Estimation of GLMM trees}

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors. Further, $X_i$ is the $N_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises a column of 1s for the intercept and a column for the observed timepoints of subject $i$. The set of observed timepoints may differ between subjects, and thus the number of rows of $X_i$ may differ between subjects. $\beta_j$ is a column vector of node-specific fixed-effects parameters, their values depending on terminal node $j$ of which observation $i$ is part. Subscript $j$ is what distinguishes the GLMM tree model from a traditional GLMM. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$ and contains (a subset of) the columns of $X_i$. Further, $b_i$ is a vector of random effects of subject $i$. We assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$. We assume Gaussian errors, with constant variance across subjects.

The parameters of a traditional GLMM can be estimated by, for example, (restricted) maximum likelihood. The GLMM tree model in Eq. \ref{eq:GLMMtree} is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:

@@

The algorithm initializes by setting $b$ to $0$, since the random effects are initially unknown. In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2. Note that the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset. The fixed effects are estimated \textit{locally}, using the observations in the current node. Convergence of the algorithm is monitored by computing the log-likelihood criterion of the mixed-effects model fitted in Step~3. Typically, this converges if the tree does not change from one iteration to the next.

@@

Studies to date indicate that initializing estimation with zero random effects yields an accurate final model \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed the impact of different initial values and found only minor impact on the final model, which became smaller with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional datasets. With longitudinal data (e.g., LGCMs), observations from the same unit tend to be more strongly correlated than in cross-sectional data; initialization with zero random effects may provide an unrealistic starting point which may be difficult to overcome in subsequent iterations. Initializing GLMM-tree estimation by estimating the random effects, assuming no grouping structure, may provide a better starting point. We thus hypothesize that for partitioning LGCMs, initializing GLMM-tree estimation with the random effects can improve subgroup recovery and predictive accuracy.

@@

The partition (or tree) in Step~1 is estimated using model-based recursive partitioning \citep[MOB; ][]{ZeilyHoth08}. The MOB algorithm cycles iteratively through the following steps:

\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\alph{enumi})}
	\item Fit the parametric model to all observations in the dataset. 
	\item Statistically test for parameter instability with respect to each of the partitioning variables.
	\item If there is some overall parameter instability, split the dataset with respect to the partitioning variable associated with the highest instability.
	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
\end{enumerate}

Parameter stability is tested in Step~(b), using the the \textit{scores} (gradient contributions) from the model fitted in Step~(a). Under mild regularity conditions, the scores have an expected value of 0. The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or if they exhibit systematic deviations when they are ordered by the values of one or more covariates. For continuous covariates (or ordered covariates with a large enough number of unique values), this involves computing the cumulative score process $W_k(t)$ with respect to each potential partitioning variable $U_k$ \citep{ZeilyHoth08}:

@@

where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in the current node, and $n$ gives the number of observations in the current node. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes estimated scores (denoted $\hat{\psi}$), with the subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $nt = 1$ for an observation with a unique minimum on the partitioning variable, and $nt = n$ for an observation with a unique maximum.

@@

When partitioning LGCMs, covariates will often not be measured at the lowest level, but at the subject level. The covariance matrix employed in Eq. \ref{eq:efp} can account for this clustering of observations within subjects \citep[e.g., ][]{ZeilyKoll20} by employing a clustered covariance matrix $\hat{J}$ in computation of the score process. \textcolor{orange}{@AZ: Thus, only computation of $\hat{J}$ is affected by cluster argument? Sscores are summed before computing $\hat{J}$, but computation of $\sum^{[nt]}_{i=1}{\hat{\psi}}_{\sigma(U_{ik}})$ is unaffected?} This resembles a GEE-type approach, where dependence between observations is accounted for in the variance structure. We thus hypothesize that for partitioning LGCMs, use of clustered covariances yields more accurate subgroup recovery and improves predictive accuracy.

\section{Study I: Performance of LM(M) trees}

To test whether initializing GLMM-tree estimation with the random effects and use of clustered covariances in computation of the parameter stability tests yields more accurate subgroup recovery when partitioning LGCMs, we performed a simulation study.

@@

\begin{figure}[!ht]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
<<fig=T, height=5, width=5, echo=F>>=
#plot(tree, ip_args = list(pval = FALSE), gp = gpar(cex = .7),
#     tp_args = list(pointcol = "white", xlab = "time",
#                    mainlab = function(id, nobs) paste("node", id)))
plot(tree, tnex = 3.5, terminal_panel = combine_panels(
    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
                        mainlab = function(id, nobs) paste("node", id)),
    panel2 = node_terminal(fig4, FUN = identity, height = 4, width = 12, id = FALSE),
    party2 = fig4), 
  gp = gpar(cex = .7), ip_args = list(pval = FALSE))
@
\label{fig:design_tree}
\end{figure}

We simulated datasets according to the subgroup structure depicted in Figure~\ref{fig:design_tree}. Specifically, four non-overlapping subgroups corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree} were created. Observations in terminal nodes 3 and 4, and in terminal nodes 6 and 7 have the same fixed intercept ($\beta_{j0}$). To facilitate comparison with the SEM tree algorithm, the same timepoints were generated for all subjects (i.e., $t = 0, 1, 2, 3, 4$). The response was computed as:\\

@@

\noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part. The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints. The value for $b_i$ was generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next. Values in $\epsilon_{i}$ were independently generated from $N(\mu = 0, \sigma = \sqrt{5})$.

@@

We employed a full factorial design, yielding $2^5 = 32$ cells.We performed 100 repetitions for each cell of the design. For generating and analysing data, we used R \citep[][version 4.1.2]{R22}.

We fitted two different LM trees to each dataset: One employing default settings (i.e., parameter stability tests performed at level I) and one where the parameter stability tests employed clustered covariance matrices (i.e., parameter stability tests performed at level 2). To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15][]{HothyZeil15}.

We fitted eight different LMM trees to each dataset: We estimated an LMM tree using default settings (i.e., parameter stability tests performed using observation-level covariances, estimation initialized with the tree structure); an LMM tree with parameter stability tests performed using cluster-level covariances; an LMM tree estimated by initializing estimation with the random effects; an LMM tree estimated by combining the latter two approaches. Each of the four approaches were applied with a specification comprising random intercepts only, as well as with a specification comprising both random intercepts and slopes, yielding eight different LMM trees. The different specifications allow for assessing the effect of mis-specification of the random-effects structure. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29][]{BateyMach15} and to fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0][]{FokkySmit18}.

@@

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. The default fitting approach overfits, and implemented $> 3$ splits in most datasets. This overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees (mean number of splits \Sexpr{round(means["lmer_c"], digits = 2)} for trees estimating only random intercepts; \Sexpr{round(means["lmer_s_c"], digits = 2)} for trees estimating random intercepts and slopes).

Initializing estimation with the random effects also effectively reduces tree size, but too strongly when random slopes are estimated (mean number of splits without random slopes \Sexpr{round(means["lmer_r"], digits = 2)}; with random slopes \Sexpr{round(means["lmer_s_r"], digits = 2)}). Heterogeneity related to partitioning variables measured at level 2 can either be captured and explained by the tree, or captured by the random effects. By initializing estimation with the random effects effects, this heterogeneity can no longer be captured by the tree; this likely worsens with more complex random-effects specification. Combining the use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding a similar and too high number of splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).

Distributions of the number of splits, separated according to the levels of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ are depicted in Figure~\ref{fig:LMM_sizes_interact} (Appendix~\ref{sec:AppendixA}). These plots show a pattern similar to Figure~\ref{fig:LMM_sizes}, with use of cluster-level covariances providing the most accurate split recovery. In addition, the plots show that increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ yield higher numbers of splits, while the effect of $\rho$ is very minor. \textcolor{orange}{I moved plots to Appendix because of their size and because they are not of main interest. Not sure if following results should be discussed here or in the Appendix:} The main deviation from this pattern of main effects is that increased variance of the random effects (increases $\sigma_{b_0}$ and/or $\sigma_{b_1}$) yields lower numbers of splits if random slopes are estimated and estimation is initialized with the random effects.

>>>>

\abstract{Growth curve models are popular tools for studying the development of a response variable within subjects over time. Between-subject heterogeneity is commonly encountered and often of specific research interest. We show how generalized linear mixed effects model (GLMM) trees can be used to identify heterogeneous subgroups in linear growth curve models. GLMM trees were originally proposed for clustered cross-sectional data; here we introduces extensions for longitudinal data. The resulting extended GLMM trees are directly applicable to growth curve models as an important special case. We evaluate the performance of GLMM trees in partitioning growth curve models on simulated datasets and existing datasets on children's development of reading, math and science abilities. We compare the performance of GLMM trees with the SEM trees, LongCART and longRPart methods. GLMM trees provide similar accuracy compared to SEM trees, and better accuracy compared to LongCART and longRPart. In addition, GLMM trees allow for analysing discrete as well as continuous time series, are less sensitive to (mis-)specification of the random-effects structure and require substantially lower computation times.}

@@

Modeling development over time is of prime interest in psychological research. For example, researchers may want to model student's reading or math development in an educational study, or symptom reduction over the course of a clinical trail. Inter-individual heterogeneity is likely encountered in such studies. For example, some (groups of) participants may show higher or lower rates of growth over time. Development over time is generally studied with mixed-effects or latent-variable models, which allow for capturing inter-individual heterogeneity and explaining this heterogeneity with covariates of a-priori known relevance \citep{NeisyMatt18}.

However, the covariates that predict or affect developmental trajectories may not be known in advance and often researchers may have a specific interest in identifying these covariates. Recursive partitioning methods (RPMs) are exploratory methods that allow for identifying relevant predictors from a potentially (very) large number of covariates. At the same time, RPMs allow for direct identification of heterogeneous subgroups. Several existing RPMs allow for subgroup detection in growth-curve models: GUIDE \citep{Loh02}, longRPart \citep{AbdoyLeBl02}, GEE-based decision trees \citep{Lee05}, longitudinal interaction trees \cite[IT; ][]{SuyMene11}, structural equation model (SEM) trees \citep{BranyOert13, ArnoyVoel21}, mixed-effects longitudinal trees \citep[MELT; ][]{EoyCho14} and LongCART \cite{KundyHare19} allow for partitioning linear growth curve models (LGCMs). Further, longRPart2 \citep{StegyJaco18} and IT-LT \citep{WeiyLiu20} methods allow for subgroup detection in non-linear growth curve models \footnote{Both IT methods have a specific aim of detecting subgroups for which the time-by-treatment interaction differs, so are not general methods for partitioning growth curve models.}.

In this paper, we propose the use of generalized linear mixed-effects model trees \citep[GLMM trees; ][]{FokkySmit18} for subgroup detection in LGCMs. GLMM trees were originally proposed for subgroup detection in clustered cross-sectional studies, where subjects are for example nested in treatment centers, classrooms and/or geographical areas. Yet, GLMM trees provide a much more general method, allowing for subgroup detection in almost any type of mixed-effects model. Compared to clustered cross-sectional data, however, longitudinal data often come with two central characteristics that may require a different estimation approach: Higher random-effects variance (i.e., higher intra-class correlations) and predictors of interest measured at higher levels (e.g., time-constant covariates). In this paper, we introduce and test two recent extensions of GLMM trees that can account for these characteristics. We focus on the specific use case of partitioning LGCMs, but methods for dealing with stronger intra-class correlations and higher-level covariates of interest are obviously of relevance beyond the realm of LGCMs.

\subsection{Recursive partitioning methods for longitudinal data}

Compared to the aforementioned RPMs for partitioning LGCMs \citep{Loh02, AbdoyLeBl02, Lee05, SuyMene11, BranyOert13, ArnoyVoel21, EoyCho14, KundyHare19}, GLMM trees differ in that they do not fit a full parametric model in each of the terminal nodes. GLMM trees estimate the fixed-effects parameters locally, using only the observations within a terminal node, and estimate the random-effects parameters globally, using all observations. Not only does this reduce computation time, it also reduces model complexity, which in turn may reduce instability and overfitting. This local-global estimation approach was first proposed by \cite{HajjyBell11} and \cite{SelaySimo12} for trees with constant fits in the terminal nodes. In effect, the approach replaces the fixed-effects part of the mixed-effects model with a decision tree with constant fits in the terminal nodes, while the random effects are estimated as usual. \cite{FokkySmit18} generalized this approach to allow for GLMs with any number of parameters in the terminal nodes, thus allowing for non-Gaussian responses and targeted detection of interaction effects.

The computational advantage of GLMM trees is highest compared to longRPart, longRPart2, IT-LT and LRT-based SEM trees. These latter methods employ an exhaustive split detection procedure: For every possible split in the current node, the full parametric model is re-estimated in the two nodes that would result from the split. To choose the optimal split, a likelihood ratio test is computed for every possible cut point. In addition to a heavy computational burden, this also introduces a selection bias towards covariates with a larger number of possible cut points \citep{ShihyTsai04, Shih04}. LongCART, MELT, GEE-based decision trees and score-based SEM trees also fit full parametric models in each of the nodes, but do not refit models for cutpoint selection; they employ the predictions or residuals from the fitted model in the current node for selecting the best split. This reduces the computational burden, and allows for separating variable and cutpoint selection, thus preventing variable selection bias. The GLMM tree algorithm is more similar to the latter group of RPMs, in that it employs a two-step approach to split detection.

\subsubsection{Current paper}

In what follows, the next section explains how GLMM trees are estimated, and how it can be adapted to better account for longitudinal data structures. In three empirical evaluations, we assess and compare performance of GLMM trees in partitioning LGCMs: In Simulation Study I, we assess performance of the proposed adaptations to GLMM trees. In Simulation Study II, we compare performance of GLMM trees with that of two other methods: SEM trees and LongCART. In the Application, we compare the performance of GLMM trees and longRPart in partitioning existing datasets on children's development of reading, math and science abilities. In the Discussion, we summarize our findings and their implications.

@@

\section{Estimation}
\label{sec:estimation}

In the GLMM tree model \citep{FokkySmit18}, expectation $\mu_i$ of outcome vector $y_i$ is modeled through a linear predictor and suitable link function:
%
\begin{eqnarray}
\label{eq:expected_value}
E[y_i | X_i] & = & \mu_i, \\
\label{eq:GLMMtree}
g(\mu_{i}) & = & X_{i} \beta_{j} + Z_{i} b_{i}
\end{eqnarray}
%
In the current paper, $g$ is the identity function and we assume a continuous response with normally distributed errors and constant variance $\sigma_\epsilon$. In principle, generalizations to other distributions within the GLM. Further, $X_i$ is the $N_i \times (p+1)$ fixed-effects design matrix for subject $i$ ($i = {1, \dots, N}$). We furter assume time is the predictor variable of interest, thus $p=1$ and $X_i$ comprises two columns: a column of 1s for the intercept and a column of observed timepoints for subject $i$. The number ($N_i$) and values of observed timepoints may differ between subjects. While in a traditional GLMM, the fixed-effects parameters are assumed constant over all observations, but in the GLMM tree model, $\beta_j$ is a column vector of node-specific fixed-effects parameters, its value depending on terminal node $j$ of which observation $i$ is part. As in a traditional GLMM, $Z_i$ is the random-effects design matrix for subject $i$, comprising a subset of columns of $X_i$ and $b_i$ is a vector of random effects for subject $i$. Finally, we assume that $b$ follows a (possibly multivariate) normal distribution with mean zero and (co)variance $\Sigma_{b}$.

The parameters of a traditional GLMM can be estimated by, for example, (restricted) maximum likelihood. The GLMM tree model is estimated by alternating between estimating the partition (i.e., subgroups or terminal nodes $j$), and estimating the random- and fixed-effects parameters, as per the following algorithm:

@@

The algorithm initializes by assuming zero random effects. In every iteration, the partition is (re-)estimated using a GLM-based recursive partition in Step~1, while the fixed- and random-effects parameters are (re-)estimated in Step~2. Thus, the random effects are not partitioned, but estimated \textit{globally}, using all observations in the dataset. The fixed effects are estimated \textit{locally}, using the observations in the current node. Convergence of the algorithm is monitored through the log-likelihood of the mixed-effects model fitted in Step~3. Typically, this converges if the tree does not change from one iteration to the next.

@@

Studies to date indicate that initializing estimation with zero random effects yields an accurate final model \citep{HajjyBell11, HajjyBell14, HajjyLaro17, SelaySimo12, FuySimo15, FokkySmit18}. \cite{SelaySimo12} assessed found only minor impact of different initialization values on the final model, which became smaller with increasing sample size. In \cite{FokkySmit18}, we found initializing estimation of GLMM trees with zero random effects performed well in cross-sectional clustered data. With longitudinal data, repeated measures on the same subjects tend to be correlated more strongly than observations nested withiin the same unit in cross-sectional data. If the random effects are in fact sizable, assuming them to be zero could provide an unrealistic starting point that is difficult to overcome in subsequent iterations. Our first hypothesis is that for partitioning LGCMs, initializing estimation with the random effects instead of the subgroup structure improves accuracy of GLMM trees.

@@

The subgroup structure in Step~1 is estimated using model-based recursive partitioning \citep[MOB; ][]{ZeilyHoth08}. The MOB algorithm cycles iteratively through the following steps:

\begin{enumerate}
\setlength\itemsep{0em}
\renewcommand{\labelenumi}{\alph{enumi})}
	\item Fit the parametric model to all observations in the dataset. 
	\item Test for parameter instability with respect to each of the partitioning variables.
	\item If there is some overall parameter instability, split the dataset with respect to the partitioning variable associated with the highest instability.
	\item Repeat Steps (a) through (c) in each of the resulting subgroups.
\end{enumerate}

Parameter stability in Step~(b) is tested using the the \textit{scores} (gradient contributions) from the model fitted in Step~(a). Under mild regularity conditions, the scores have an expected value of 0. The parameter stability tests evaluate whether the scores fluctuate randomly around their mean of 0, or exhibit systematic deviations when ordered by the values of a covariate. For continuous covariates (or ordered covariates with a large enough number of unique values), this involves computing the cumulative score process $W_k(t)$ with respect to each potential partitioning variable $U_k$ \citep{ZeilyHoth08}:

@@

where $\hat{J}$ is a suitable estimate of the covariance matrix of the parametric model fitted in Step~(a), and $n$ gives the number of observations in the current node. Further, $\hat{\psi}_{\sigma(U_{ik})}$ denotes estimated scores (denoted $\hat{\psi}$), with the subscript {$\sigma(U_{ik})$} denoting their ordering by the values of partitioning variable $U_k$. Note that $0 \leq t \leq 1$, thus $nt = 1$ for an observation with a unique minimum on the partitioning variable, and $nt = n$ for an observation with a unique maximum.

\textcolor{red}{Perhaps replace $n$ with $N_j$? This is more line with notation of GLMM tree model, but might be confusing with cov mat $J$?}

@@

When partitioning longitudinal data, covariates will often be measured at the subject level. This clustering of observations within subjects can be accounted for in computing a so-called \textit{clustered} covariance matrix $\hat{J}$ \citep[e.g., ][]{ZeilyKoll20}. This resembles a GEE-type approach with an independence correlation structure. Our second hypothesis is that for partitioning LGCMs, use of clustered covariances improves accuracy of GLMM trees.

\section{Simulation Study I}

We performed a simulation study to compare the performance GLM(M) trees estimated by initializing estimation with the subgroup structure versus the random effects, and employing observation- versus cluster-level covariances for the parameter stability tests.

@@

\begin{figure}[!t]
\caption{Design of the subgroup-specific fixed effects in the simulation.}
<<fig=T, height=5, width=5, echo=F>>=
#plot(tree, ip_args = list(pval = FALSE), gp = gpar(cex = .7),
#     tp_args = list(pointcol = "white", xlab = "time",
#                    mainlab = function(id, nobs) paste("node", id)))
plot(tree, tnex = 3.5, terminal_panel = combine_panels(
    panel1 = node_bivplot(tree, xlab = "time", pointcol = "white",
                        mainlab = function(id, nobs) paste("node", id)),
    panel2 = node_terminal(fig4, FUN = identity, height = 4, width = 12, id = FALSE),
    party2 = fig4), 
  gp = gpar(cex = .7), ip_args = list(pval = FALSE))
@
\label{fig:design_tree}
\end{figure}

We simulated data according to the subgroup structure depicted in Figure~\ref{fig:design_tree}. Every dataset comprised four non-overlapping subgroups, corresponding to the terminal nodes of the tree in Figure~\ref{fig:design_tree}. The response was computed as:\\

@@

\noindent where $\beta_j$ corresponds to the fixed effects in terminal node $j$ of which subject $i$ is part. The bottom panels of Figure~\ref{fig:design_tree} present the subgroup-specific values of the fixed intercept and slope ($\beta_{j0}$ and $\beta_{j1}$, respectively). The fixed- and random-effects design matrices $X_i$ and $Z_i$ are identical, each comprising two columns: a vector of 1s for the intercept, and a vector of timepoints. The same set of timepoints was generated for all subjects: ${0, 1, 2, 3, 4}$ \footnote{Although GLMM trees do not require the same set of timepoints for each subject, this design allowed for comparison with SEM trees.}. Values of $b_i$ were generated from a multivariate normal distribution with mean zero and a $2 \times 2$ diagonal covariance matrix $\Sigma$, the diagonal entries determined by the level of the data-generating design described next. Values in $\epsilon_{i}$ were independently generated from $N(\mu = 0, \sigma = \sqrt{5})$.

@@

We employed a full factorial design yielding $2^5 = 32$ cells and performed 100 repetitions per cell. We used R \citep[version 4.1.2; ][]{R22} to generate and analyze all data.

\subsubsection{Partitioning approaches}

We fitted ten trees to each dataset:

\begin{itemize}
  \item Two linear model (LM) trees: 
  \begin{itemize}
    \item A tree employing default settings, where the parameter stability tests use observation-level covariances.
    \item A tree where the parameter stability tests use clustered covariances. 
  \end{itemize}
  \item Eight linear mixed model (LMM) trees: 
  \begin{itemize}
    \item Two trees fitted using default settings, where parameter stability tests use observation-level covariances and estimation is initialized with the tree structure. 
    \item Two trees where the parameter stability tests use cluster-level covariances.
    \item Two trees where estimation is initialized with the random effects. 
    \item Two trees where the parameter stability tests use cluster-level covariances and estimation is initialized with the random effects.
  \end{itemize}
\end{itemize}

Each of the two LMM trees were fitted with different specifications of the random effects: One of the LMM trees estimated random intercepts only, one of the LMM trees estimated both random intercepts and slopes. Because variances of random intercepts and slopes were $> 0$ in all cells of the design, the three different random-effects specifications (LM trees assume 0 random effects) allow for assessing the effect of mis-specifying the random effects. To fit LM trees, we used package \textbf{partykit} \citep[version 1.2-15; ][]{HothyZeil15}. To fit LMMs, we used package \textbf{lme4} \cite[version 1.1-29; ][]{BateyMach15}. To fit LMM trees we used package \textbf{glmertree} \citep[version 0.2-0; ][]{FokkySmit18}.

@@

Figure~\ref{fig:LMM_sizes} depicts the number of splits implemented by each partitioning approach. Irrespective of the random-effect specification, the default fitting approaches overfit, implementing $> 3$ splits in a large majority of datasets. Irrespective of the random-effects specification, the overfitting is successfully mitigated by the use of cluster-level covariances in the parameter stability tests (mean number of splits \Sexpr{round(M_cluster, digits = 2)}), which appears most effective for LMM trees: the mean number of splits was \Sexpr{round(means["lmer_c"], digits = 2)} for LMM trees with random intercepts only and \Sexpr{round(means["lmer_s_c"], digits = 2)} for LMM trees estimating both random intercepts and slopes.

Initializing estimation with the random effects was also effective in reducing tree size, bput the effect depended on the random-effects specification. When random slopes were estimated in addition to random intercepts, too few splits were implemented on average: the mean number of splits was \Sexpr{round(means["lmer_r"], digits = 2)} for LMM trees with random intercepts only, and \Sexpr{round(means["lmer_s_r"], digits = 2)}) for LMM trees estimating both random intercepts and slopes. Combined use of random-effects initialization and cluster-level covariances does not appear beneficial, yielding too many splits, irrespective of the random-effects specification (mean number of splits \Sexpr{round(M_both, digits = 2)}).

Distributions of the number of splits, separated according to the levels of the data-generating design are depicted in Figure~\ref{fig:LMM_sizes_interact} (Appendix~\ref{sec:AppendixA}). The plots indicate little interaction between data-generating parameters and model-fitting approached. Figure~\ref{fig:LMM_sizes_interact} shows a pattern very similar to Figure~\ref{fig:LMM_sizes}, where the use of cluster-level covariances provides the most accurate split recovery. Main effects of the data-generating design were as expected: Increasing values of $\sigma_{b_0}$, $N$, $p$ and $\sigma_{b_1}$ yield higher numbers of splits. The effect of $\rho$ was minor. The only evidence of an interaction between data-generating parameters and model-fitting approaches is that increased variance of the random effects (higher $\sigma_{b_0}$ and/or $\sigma_{b_1}$) yields lower numbers of splits if random slopes are estimated and estimation is initialized with the random effects. Thus, with stronger random effects and a more complex random-effects specification, differences in fixed effects (which should be captured by the tree) tend to be captured by the random effects when estimation is intialized with the random effects.
